sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-27 doc/paper.Rmd
Add index to pcod fig

```{r pkg-checks, include=FALSE, echo=FALSE}
if (!require("rnaturalearthhires", quietly = TRUE)) {
  stop("Please install 'rnaturalearthhires'.\n",
  "`remotes::install_github('ropensci/rnaturalearthhires')`")
}
if (!require("ggsidekick", quietly = TRUE)) {
  stop("Please install 'ggsidekick'.\n",
  "`remotes::install_github('seananderson/ggsidekick')`")
}
```

@@

```{r, echo=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

@@

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

@@

<!-- tidy(fit, conf.int = TRUE) -->
<!-- tidy(fit, effects = 'ran_pars', conf.int = TRUE) -->

@@

```{r, echo=FALSE, eval=TRUE}
p <- predict(fit, newdata = subset(qcs_grid, year == 2003))
```

```{r, eval=FALSE}
p3 <- predict(fit, newdata = subset(qcs_grid, year == 2003), sims = 500L)
se <- apply(p3, 1, sd)
p$se <- se
g_pred_se <- ggplot(p, aes(X, Y, fill = se)) +
  geom_raster() +
  scale_fill_viridis_c(trans = "log10", option = "C") +
  coord_fixed() +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt"))
```

```{r, warning=FALSE, message=FALSE}
library(sf)
# devtools::install_github("ropensci/rnaturalearthhires") # install if needed
map_data <- rnaturalearth::ne_countries(
  scale = "large",
  returnclass = "sf", country = "canada")
# Crop the polygon for plotting and efficiency:
# st_bbox(map_data) # find the rough coordinates
bc_coast <- suppressWarnings(suppressMessages(
  st_crop(map_data,
    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
utm_zone9 <- 3156
bc_coast_proj <- sf::st_transform(bc_coast, crs = utm_zone9)
# ggplot(bc_coast_proj) + geom_sf()
```

```{r}
g_pred <- ggplot(bc_coast_proj) +
  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = exp(est))) +
  coord_sf(datum = st_crs(utm_zone9)) +
  theme(panel.grid.major = element_line(colour = "grey90")) +
  scale_x_continuous(
    limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_y_continuous(
    limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_fill_viridis_c(trans = "sqrt", option = "C") +
  # scale_fill_viridis_c(trans = "sqrt", option = "B", begin = 0.3) +
  labs(fill = "Predicted\ndensity (units)", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"),
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
  )
```

```{r, echo=FALSE}
blues <- RColorBrewer::brewer.pal(5, "Blues")
reds <- RColorBrewer::brewer.pal(5, "Reds")
g_mesh <- ggplot(pcod, aes(X, Y)) +
  inlabru::gg(mesh$mesh, edge.color = "grey60") +
  geom_point(alpha = 0.6, mapping = aes(size = density), colour = blues[4]) +
  coord_fixed() +
  labs(x = "UTMs E (km)", y = "UTMs N (km)") +
  scale_size_area(max_size = 10) +
  theme(panel.grid.major = element_line(colour = "grey90")) +
  theme(legend.position = "none")
```

```{r}
nd <- data.frame(
  depth =
    seq(min(pcod$depth), max(pcod$depth), length.out = 300)
)
p_smooth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA)
g_smooth <- ggplot(
  p_smooth,
  aes(depth, exp(est),
    ymin = exp(est - 1.96 * est_se),
    ymax = exp(est + 1.96 * est_se)
  )
) +
  geom_ribbon(fill = blues[2]) +
  geom_line(colour = blues[5], lwd = 1) +
  labs(x = "Depth (m)", y = "Density (units)") +
  xlim(min(pcod$depth), 300) +
  coord_fixed(expand = FALSE, ratio = 1.28) +
  theme(panel.grid.major = element_line(colour = "grey90"))
```

```{r fit-pcod-st, warning=FALSE, message=FALSE, cache=TRUE, eval=FALSE, echo=FALSE}
fit2 <- sdmTMB(
  density ~ 0 + as.factor(year),
  family = tweedie(link = "log"), data = sdmTMB::pcod, mesh = mesh,
  time = "year", spatial = "on", spatiotemporal = "iid", silent = TRUE
)
p2 <- predict(fit2, newdata = qcs_grid, sims = 500L)
ind <- get_index_sims(p2)
```

```{r plot-ind, eval=FALSE}
g_ind <- ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) + 
  geom_ribbon(fill = blues[2]) +
  geom_line(colour = blues[5], lwd = 1) +
  ylim(0, max(ind$upr)) +
  coord_fixed(expand = FALSE, ratio = 0.3) +
  labs(x = "Year", y = "Biomass (units)")
```

@@

```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", fig.width=9, out.width="5.2in", fig.asp=1, fig.align='center'}
g_omega <- ggplot(bc_coast_proj) +
  geom_sf(fill = NA, colour = "gray30", lwd = 0.3) +
  geom_raster(data = p, aes(x = X * 1000, y = Y * 1000, fill = omega_s)) +
  coord_sf(datum = st_crs(utm_zone9)) +
  scale_x_continuous(
    limits = c(230957.7 + 105000, 1157991 - 570000),
    labels = function(x) paste0(x / 1000)
  ) +
  scale_y_continuous(
    limits = c(5366427 + 270000, 6353456 - 513000),
    labels = function(x) paste0(x / 1000)
  ) +
  # scale_fill_gradient2(high = reds[4], low = scales::muted("purple")) +
  scale_fill_gradient2(high = reds[4], low = blues[4]) +
  labs(fill = "Spatial random\nfield deviations", x = "UTMs E (km)", y = "UTMs N (km)") +
  theme(
    # panel.background = element_rect(fill = "slategray1", colour= "white"),
    legend.position = c(0.2, 0.2), legend.key.height = unit(10, "pt")
  ) +
  theme(panel.grid.major = element_line(colour = "grey90"))
cowplot::plot_grid(g_mesh, g_smooth, g_omega, g_pred, labels = "auto", align = "hv")
```

```{r, eval=FALSE, echo=TRUE}
fit_spatiotemporal <- sdmTMB(
  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
  time = "year", spatial = "off", spatiotemporal = "ar1"
)
```

We could generate a standardized index from that fit with:

```{r, eval=FALSE, echo=TRUE}
p <- predict(fit_spatiotemporal, newdata = qcs_grid, return_tmb_object = TRUE)
ind <- get_index(p)
```

As a third example, we could model presence/absence of Pacific cod (`present`) with a time-varying quadratic effect of depth, an independent mean each year, a spatial random field, and IID spatiotemporal fields:

```{r, eval=TRUE, echo=FALSE}
# restore 'present' column
pcod <- sdmTMB::pcod
```

```{r, eval=FALSE, echo=TRUE}
fit_spatiotemporal_tv <- sdmTMB(
  present ~ 0 + as.factor(year), 
  time_varying = ~ 0 + poly(depth, 2), 
  family = binomial(), data = pcod, mesh = mesh,
  time = "year", spatial = "on", spatiotemporal = "iid"
)
```

@@

```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="3.9in", fig.align='center'}
owl <- here::here("figs", "owl_nao_effect.pdf")
if (!file.exists(owl)) {
  download.file("https://www.dropbox.com/s/fz1iw0a0diohyxc/owl_nao_effect.pdf?dl=1", 
    destfile = owl)
}
knitr::include_graphics(owl)
```

>>>>

```{r pkg-checks, include=FALSE, echo=FALSE}
# if (!require("rnaturalearthhires", quietly = TRUE)) {
#   stop("Please install 'rnaturalearthhires'.\n",
#   "`remotes::install_github('ropensci/rnaturalearthhires')`")
# }
# if (!require("ggsidekick", quietly = TRUE)) {
#   stop("Please install 'ggsidekick'.\n",
#   "`remotes::install_github('seananderson/ggsidekick')`")
# }
```

@@

```{r, echo=TRUE, eval=FALSE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

@@

```{r pcod-eg1-fit, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod,
  family = tweedie(link = "log"),
  mesh = mesh,
  spatial = "on"
)
```

@@

@@

```{r fit-tv, eval=FALSE, echo=TRUE}
fit_spatiotemporal <- sdmTMB(
  density ~ s(depth), family = tweedie(link = "log"), data = pcod, mesh = mesh,
  time = "year", spatial = "off", spatiotemporal = "ar1"
)
```

We could generate a standardized index from that fit on a 2x2 km grid:

```{r, eval=FALSE, echo=TRUE}
p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
ind <- get_index(p)
```

@@

```{r pcod-fig, fig.cap="(ref:pcod-fig-cap)", out.width="5.2in", fig.align='center'}
pcodfig <- here::here("figs", "pcod-fig.pdf")
if (!file.exists(pcodfig)) {
  download.file("https://www.dropbox.com/s/s8kl4u49p27qqhy/pcod-fig.pdf?dl=1", 
    destfile = pcodfig)
}
knitr::include_graphics(pcodfig)
```

@@

```{r owl-nao, fig.cap="(ref:owl-nao-fig)", out.width="4.1in", fig.align='center'}
owl <- here::here("figs", "owl_nao_effect.pdf")
if (!file.exists(owl)) {
  download.file("https://www.dropbox.com/s/fz1iw0a0diohyxc/owl_nao_effect.pdf?dl=1", 
    destfile = owl)
}
knitr::include_graphics(owl)
```
